A Formal Theory of Bidirectional Typing for Simple Types

### 2023-05-25 tex/BiSig.tex
[ fix ] Use mode-indexing judgement instead of the conventional mutually defined judgements

\section{Bidirectional simply typed \texorpdfstring{$\lambda$}{lambda}-calculus}\label{sec:bi-stlc}
\begin{figure}
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma \vdash \isTerm{t} \Rightarrow A}
    \quad\text{and}\quad \boxed{\Theta \mid \Gamma \vdash \isTerm{t} \Leftarrow A}
    \quad\text{where $\Theta \vdash \Gamma$ and $\Theta \vdash A$} 
    \\
    \inferrule{(x : A) \in \Gamma}{\Theta \mid \Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
    \and
    \inferrule{\Theta \mid \Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Theta \mid \Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
    \and
    \inferrule{\Theta \mid \Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash \isTerm{t\uparrow} \Leftarrow A}\;(\text{sub})
    \and
    \inferrule{\Theta \mid \Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Theta \mid \Gamma \vdash \isTerm{u} \Leftarrow A}{\Theta \mid \Gamma \vdash \isTerm{t\;u} \Rightarrow B}
    \and
    \inferrule{\Theta \mid \Gamma, \isTerm{x} : A \vdash \isTerm{t} \Rightarrow B}{\Theta \mid \Gamma \vdash \isTerm{\lam{x}t} \Rightarrow A \bto B}
  \end{mathpar}
  \caption{The extrinsic style of bidirectional simply typed $\lambda$-calculus with type variables}
  \label{fig:bi-stlc}
\end{figure}
\begin{remark}
  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
\end{remark}

@@

\section{A Definition of Simple Type Theories}\label{sec:simple-type-theories}
\subsection{Simple Types}
\begin{definition}
  A \emph{(simple type) signature} $\Sigma$ consists of a set $I$ and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
  An inhabitant $i : I$ is meant to represent the \emph{$i$-th operation} $\tyOp_i$ and its \emph{number} $ \arity(i)$ of arguments.

@@

Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A}$ constructed with only the rule $(\var)$ and the rule scheme $(\termOp_o)$ for every operation $o$ in $\Omega$ displayed in \Cref{fig:intrinsic-typing}.

The substitution $\rho\colon \Xi \to \Theta$ is used to instantiate explicitly variables in the local context $\Xi$ with global type variables in $\Theta$.
Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\termOp_o$ to ensure that all types are will-formed under $\Theta$.

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A} \quad \text{where $\Theta \vdash_{\Sigma} \Gamma$ and $\Theta \vdash_{\Sigma} A$} \\
    \inferrule{A \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
    \and
    \inferrule*{\rho : \Sub{\Xi}{\Theta} \\ \Theta \mid \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
    {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\termOp_o)
  \end{mathpar}
  for $o\colon \Xi \vdash [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$
  \caption{Intrinsic typing rules}
  \label{fig:intrinsic-typing}
\end{figure}
\subsection{Raw Terms and Extrinsically Typed Terms?}
\begin{LTannote}{no extrinsic typing}
Argue that we don't need extrinsically typed terms but raw terms only as the output of some parser.
\end{LTannote}

\begin{figure}
  \centering
  \begin{mathpar}
    \boxed{\Theta \vdash_{\Sigma, \Omega} t}
  \end{mathpar}

@@

\begin{definition}
  For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with a function
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \UU} \left(\Cxt_{\Sigma}(\Xi) \times \Mode \times \Type_{\Sigma}(\Xi)\right)^* \times \Mode \times \Type_{\Sigma}(\Xi).
  \]
  Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\termOp_o$ in a bidirectional system, i.e.
  an arity is a $4$-tuple
  \[
    \left(\Xi, \left[\left(\Delta_1, \leftrightarrows_1, A_1\right), \ldots, \left(\Delta_{n}, \leftrightarrows_n, A_{n}\right) \right], \leftrightarrows, A\right)
  \]
  where ${\leftrightarrows_i}$ and ${\leftrightarrows}$ are of type $\Mode$ indicating the modes of a construct and its arguments respectively.
  For brevity, we write $o \colon \Xi \vdash [\Delta_1]A_{1}^{\leftrightarrows_1}, \ldots, [\Delta_{n}] A^{\leftrightarrows_n}_{n} \to A^{\leftrightarrows}$ to indicate an operation $o$ with its arity. 
\end{definition}

\begin{example}
  With the type signature $\Sigma_{\bto}$, the term signature $\Lambda_{\bto}^{\leftrightarrows}$ for bidirectional simply typed $\lambda$-calculus introduced in \Cref{sec:bi-stlc} can be specified by operations 
  \begin{align*}
    \mathsf{app}\colon A, B \vdash (A \bto B)^{\Rightarrow}, A^{\Leftarrow} \to B^{\Rightarrow} &&
    \mathsf{abs}\colon A , B \vdash [A]B^{\Leftarrow} \to (A \bto B)^{\Leftarrow}
  \end{align*}
  like those in \Cref{ex:STLC-sig} but augmented with the mode information.
\end{example}

@@

\begin{figure}
  \centering
  \begin{mathpar}
    \boxed{\Theta \vdash_{\Sigma, \Omega} t \Rightarrow}
    \quad\text{and}\quad
    \boxed{\Theta \vdash_{\Sigma, \Omega} t \Leftarrow}
  \end{mathpar}

@@

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A}\quad\text{and}\quad\boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} 
    \quad \text{where $\Theta \vdash_{\Sigma} \Gamma$, $\Theta \vdash_{\Sigma} A$, and $\Theta \vdash_{\Sigma, \Omega} t \leftrightarrows$} 
    \\
    \inferrule{(x : A) \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{x} \mathrel{\color{red}\Rightarrow} A}\;(\var^{\Rightarrow})
    \and
    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow B \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t : B}\Rightarrow A}\;(\textcolor{red}{anno})
    \and
    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} \Leftarrow A}\;(\textcolor{red}{sub})
    \and
    \inferrule{\rho\colon \Sub{\Xi}{\Theta} \\
      \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} \mathrel{\color{red}\leftrightarrows_{1}} \simsub{A_{1}}{\rho} \\
      \cdots \\
    \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} \mathrel{\color{red}\leftrightarrows_{n}} \simsub{A_{n}}{\rho}}
    {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{\termOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} \mathrel{\color{red}\leftrightarrows} \simsub{A}{\rho}} \; (\termOp^{\leftrightarrows}_o)
  \end{mathpar}
  for $o \colon \Xi \vdash [\Delta_1]A_{1}^{\leftrightarrows_1}, \ldots, [\Delta_{n}] A^{\leftrightarrows_n}_{n} \to A^{\leftrightarrows}$ in $\Omega$
  \caption{Bidirectional extrinsic typing rules}
\end{figure}

@@

\subsection{Binding Signatures vs.\ their Bidirectional Variants}
\subsubsection{Soundness}
\subsubsection{Completeness (Annotatability)}
\subsection{Mode Correctness}

@@

\section{Formal Implementation and Examples}

>>>>

\section{Bidirectional simply typed \texorpdfstring{$\lambda$}{lambda}-calculus}\label{sec:bi-stlc}
\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma \vdash \isTerm{t} \Rightarrow A}
    \quad\text{and}\quad \boxed{\Theta \mid \Gamma \vdash \isTerm{t} \Leftarrow A}
    \quad\text{where $\Theta \vdash \Gamma$ and $\Theta \vdash A$} 
    \\
    \inferrule{(x : A) \in \Gamma}{\Theta \mid \Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
    \and
    \inferrule{\Theta \mid \Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Theta \mid \Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
    \and
    \inferrule{\Theta \mid \Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Theta \mid \Gamma \vdash \isTerm{t\uparrow} \Leftarrow A}\;(\text{sub})
    \and
    \inferrule{\Theta \mid \Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Theta \mid \Gamma \vdash \isTerm{u} \Leftarrow A}{\Theta \mid \Gamma \vdash \isTerm{t\;u} \Rightarrow B}
    \and
    \inferrule{\Theta \mid \Gamma, \isTerm{x} : A \vdash \isTerm{t} \Rightarrow B}{\Theta \mid \Gamma \vdash \isTerm{\lam{x}t} \Rightarrow A \bto B}
  \end{mathpar}
  \caption{The extrinsic style of bidirectional simply typed $\lambda$-calculus with type variables}
  \label{fig:bi-stlc}
\end{figure}
\begin{remark}
  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
\end{remark}

@@

\section{Definitions of Simple Type Theories}\label{sec:simple-type-theories}
\subsection{Simple Types}
\begin{definition}
  A \emph{(simple type) signature} $\Sigma$ consists of a set $I$ and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
  An inhabitant $i : I$ is meant to represent the \emph{$i$-th operation} $\tyOp_i$ and its \emph{number} $ \arity(i)$ of arguments.

@@

Intrinsically typed terms for a type signature $\Sigma$ and a term signature $\Omega$ are nothing more than derivations of the intrinsic typing judgement $\boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A}$ constructed with only the rule $(\var)$ and the rule scheme $(\termOp)$ displayed in \Cref{fig:intrinsic-typing}.

The substitution $\rho\colon \Sub{\Xi}{\Theta}$ is used to instantiate variables in the local context $\Xi$ with global type variables in $\Theta$.
Accordingly, $\rho$ has to be applied to types that appear in the arity to construct a term by $\termOp_o$ to ensure all types are will-formed under $\Theta$.

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A} \quad \text{where $\Theta \vdash_{\Sigma} \Gamma$ and $\Theta \vdash_{\Sigma} A$} \\
    \inferrule{A \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
    \and
    \inferrule*{\rho : \Sub{\Xi}{\Theta} \\ \Theta \mid \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \\ \cdots \\ \Theta \mid \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
    {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\termOp)
  \end{mathpar}
  for $o\colon \Xi \vdash [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$
  \caption{Intrinsic typing rules}
  \label{fig:intrinsic-typing}
\end{figure}
\subsection{Raw Terms and Extrinsically Typed Terms?}
\begin{LTannote}{no extrinsic typing}
Argue that we don't need extrinsically typed terms but raw terms only as the output of some parser.
\end{LTannote}

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \vdash_{\Sigma, \Omega} t}
  \end{mathpar}

@@

\begin{definition}
  For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with a function
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \UU} \left(\Cxt_{\Sigma}(\Xi) \times \Mode \times \Type_{\Sigma}(\Xi)\right)^* \times \Mode \times \Type_{\Sigma}(\Xi).
  \]
  where $\Mode$ consists of $\chk$ for checking and $\syn$ for synthesis.
  Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\termOp_o$ in a bidirectional system, i.e.
  an arity is a $4$-tuple
  \[
    \left(\Xi, \left[\left(\Delta_1, d_1, A_1\right), \ldots, \left(\Delta_{n}, d_n, A_{n}\right) \right], d, A\right)
  \]
  where $d$ and $d_i$'s indicate the modes of a construct and its arguments respectively.
  For brevity, we write $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ to indicate an operation $o$ with its arity. 
\end{definition}

\begin{example}
  The term signature $\Lambda_{\bto}^{\leftrightarrows}$ for bidirectional simply typed $\lambda$-calculus introduced in \Cref{sec:bi-stlc} can be specified by operations 
  \begin{align*}
    \mathsf{app}\colon A, B \vdash (A \bto B)^{\Rightarrow}, A^{\Leftarrow} \to B^{\Rightarrow} &&
    \mathsf{abs}\colon A , B \vdash [A]B^{\Leftarrow} \to (A \bto B)^{\Leftarrow}
  \end{align*}
  extending $\Lambda_{\bto}$ (\Cref{ex:STLC-sig}) with the mode information.
\end{example}

@@

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta \vdash_{\Sigma, \Omega} t^d}
  \end{mathpar}

@@

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{d}}
    \quad \text{where $d$ is either $\chk$ or $\syn$, $\Theta \vdash_{\Sigma} \Gamma$, $\Theta \vdash_{\Sigma} A$, and $\Theta \vdash_{\Sigma, \Omega} t^d$} 
     \\
    \inferrule{(x : A) \in \Gamma}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A^\syn}\;(\var)
    \and
    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^\chk \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t : B}:  A^\syn}\;(\text{anno})
    \and
    \inferrule{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\syn} \\ A = B}{\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} : A^\chk}\;(\text{sub})
    \and
  \end{mathpar}
  \begin{mathpar}
    \inferrule{\rho\colon \Sub{\Xi}{\Theta} \\
      \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} : \simsub{A_{1}}{\rho}^{\isDir{d_1}} \\
      \cdots \\
    \Theta \mid \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}^{\isDir{d_n}}}
  {\Theta \mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{\termOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\termOp)
    \\ \text{for $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
  \end{mathpar}
  Abbreviations: $\boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\syn}$ and $\boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Theta\mid \Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\chk}$
  \caption{Bidirectional extrinsic typing rules}
\end{figure}

@@

\subsection{Between Binding Signatures and their Bidirectional Variants}
\subsubsection{Soundness}
\subsubsection{Completeness (Annotatability)}
\subsection{Mode Correctness}

@@

\section{Formal Implementation and Examples}
\begin{LTannote}{de Bruijn}
  In our Agda formalisation, we use de Bruijn indices to represent variables, so we don't use an arbitrary type $\Theta$ but the number of variables as the index of typing judgements.
\end{LTannote}
